The title

### 2016-10-11 analysis/paper/paper.Rmd
more background text

### First archaeological exploration and the Movius Line

Since the 19th century, archaeological interest was focussed on the central part of the Irrawaddy River in central Myanmar. The first artefacts were discovered by Morris in 1930 in Upper Myanmar later assigned to the Anyathian culture from the Lower Palaeolithic in Southeast Asia [@movius1948lowerand]. Anyathian artefacts are mainly found in the secondary deposits of the Pleistocene terrace gravels of the Irrawaddy River, generally containing only few handaxes [@movius1948lowerand]. The occupation of the Irrawaddy River terraces during the Pleistocene seems to be limited to the east bank of the River and the central part of Myanmar because the finds concentrate in the region around Chauk, Nyaung-U and Yenanyaung [@movius1944early; @deTerra1943].

Probably due to raw material constraints (fossil wood and silicified tuff), the three phases of the Early Anyathian are dominated by a high degree of uniformity of the stone tools [@movius1948lowerand]. Fossil wood is difficult to flake because it easily flakes parallel to the natural fibres of the wood. Most of the artefacts made from fossil wood are worked on only one plane; multiple flake scars are rare. Silicified tuff is more suitable for stone tool production due to its fine-grained homogenous texture and conchodial fractures, although it can be very porous. In total, 483 implements were found (261 from fossil wood, 220 from silicified tuff and 2 from quartzite). Implements made from fossil wood include mostly handaxes, whereas stone tools made of silicified tuff are dominantly choppers [@deTerra1943; @movius1948lowerand].

In addition, 23 Early Anyathian artefacts were found in situ in the lateritic gravel and gravel deposits of terrace 1 near Chauk on the Hill of Chinaungma. Movius [-@deTerra1943] assigned these implements to the first phase of the Early Anyathian and the beginning of the Middle Pleistocene as the oldest human artefacts found in the Irrawaddy Valley. Implements of phase 2 of the Early Anyathian show either a ferruginous crust, developping during an interpluvial period, or are heavily rolled, suggesting these implements have been derived from older deposits. Most of the Artefacts from the third phase of the Early Anyathian are very heavily rolled and are associated with the early Upper Pleistocene when the gravels were deposited [@deTerra1943].

Comparing the more complex technologies found in Africa, Europe, the Levant and SW Asia and the simpler chopper/chopping tool technology of Southeast/East Asia, Movius proposed the concept of the Movius Line separating the Lower Palaeolithic into two cultures - the crude or non- stone tool-making cultures in E and SE Asia (e.g. Imjin/Hantan River Basin, Korea and Luonan Basin and Longyadong Cave, China) and northern China (e.g. Zhoukoudian) and the more complex technologies in Africa, Europe and SW Asia (e.g. Olduvai George Beds, Tanzania and Olorgesailie, Kenya) [@brumm2012biface; @norton2006middle; @gao2014late; @dennell2014hallam].

### The Movius Line in modern quaternary science

In modern quaternary science, the Movius Line and its implications on human evolution in Southeast Asia are highly debated [e.g. @brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius]. Hutterer [-@hutterer1977reinterpreting] states only questionable geological associations between artefacts and the River terraces have been made and assumptions regarding the chronology of artefacts were mostly based on the degree of rolling [@dennell2014hallam]. Norton et al. [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].

Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle]. These stone tools show a degree of standardization [@norton2006middle], included handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface] and are associated with hominins during the Pleistocene [@petraglia2008large]. Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface]. The definition of the Acheulean is not always straightforward and bifaces have been produced in different parts of the world over several 100 ka [@dennell2016life].

Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology east of the Movius Line are recently more investigated, including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius].

The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times [@dennell2016life]. There may have been no need for intensive stone tool production, or a different material was used at some point. This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius].

@@

Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g. water, food, shelter) [@lycett2010movius]. However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions [@bar2001africa; @field2005assessment; @lycett2010movius; @schick1994movius]. 
Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al. [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia. This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single] matching with some Middle Pleistocene archaeological sites in Africa, India (e.g. Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions dated to 56-69ka. A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources. However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes.

@@

Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution [@dennell2014hallam; @dennell2016life]. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, connections to other Middle Pleistocene Acheulean assemblages in Southwest Asia are difficult to demonstrate [@dennell2016life; @dennell2014hallam].

>>>>

- Movius and contemporaries

# Dennell is a major commentator recently
Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].

In contrast to Dennell’s views, Norton et al. [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].

Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.

The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times. Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].

Raw Material

@@

Environmental Changes

Whilst colonizing East Asia, early modern humans encountered barriers such as mountain ranges, river deltas, oceans and deserts that provide a range of resources (e.g. water, food, shelter) [@lycett2010movius]. However, the environment during the Middle Pleistocene was challenging in terms of changing climatic conditions and accompanying biogeographic transitions [@bar2001africa; @field2005assessment; @lycett2010movius; @schick1994movius].

There may have been no need for intensive stone tool production, or a different material was used at some point. This theory is known as the “bamboo hypothesis” [@field2005assessment; @lycett2010movius; @schick1994movius] and states that early modern humans used bamboo instead of stones to make tools [@schick1994movius] leaving behind no archaeological record [@lycett2010movius]. However, even though bamboo cutting marks on bones can be distinguished from other cutting marks, there have been no findings of such so far [@lycett2010movius].

Dispersal routes

The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists. Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al. [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia. This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single]. This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g. Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions. Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal. A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources. However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.

Demographic and social transmission

@@

[Figure demographic and social transmission?]

Definition of “Acheulean”

Acheulean bifacial tools are the earliest known artefacts from Africa dating to 1.76 Ma respectively to 1.7- 1.6 Ma [@brumm2012biface; @lycett2010movius; @norton2006middle]. In contrary to Mode I technology, these stone tools show a degree of standardization [@norton2006middle] and were the main tools used by hominins during the Pleistocene [@petraglia2008large]. The tools include handaxes, cleavers, picks, knives, lanceolates and unifaces [@brumm2012biface]. Used until 100 ka ago, Acheulean tools are commonly interpreted as butchery tools, although other functions are possible as well [@brumm2012biface]. The definition of the Acheulean is not always straightforward, i.e. the assignment of finds can be influenced by the finding location east or west of the Movius Line QUELLE. Also, bifaces are not suitable as a time marker for specific cultures as they have been produced in different parts of the world over several 100 ka [@dennell2016life].
